---
title: "Probability and statistics"
author: "Jonas Björnerstedt"
subtitle: "Statistics"
date: '`r Sys.Date()`'
output:
  ioslides_presentation:
    css: slides.css
  beamer_presentation:
    theme: default
runtime: shiny
...

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(knitr)
library(shiny)
library(grid)
library(gridExtra)
set.seed(423)
```

## Lecture Content

- Chapter 3
- More about R 

Estimation in increasing complexity is presented in chapters:

- Ch 3. Estimation of population mean of $Y$
    - Today's lectsure - the basic intuition 
- Ch 4. Estimation of relationship between $X$ and $Y$
    - Univariate regression
- Ch 6. Estimation of relationship between $X_1,..,X_k$ and $Y$
    - Multivariate regression
        
## Statistics

- _Probability theory_ - \Given probability distributions what can we say about draws? 
- _Statistics_ - What can we infer about population distribution $P(Y)$ from sample?

- _Estimator_ 
    - function of the data sample
        - a random variable!!
    - an estimator is a _statistic_
    - Often intended to describe a property of the population
        - Ex: expected value $E(Y)$ or variance $Var(Y)$

## Estimating the expected value

- Estimate the expected value of random variable $Y$
- Assume that we can take a sample of size $n$ of $Y$
$$Y_1,Y_2,Y_3,...Y_n$$
    - Independent and Identically Distributed (IID) draws
- Consider two estimators of the population mean $\mu_Y$
    1. Sample mean: $\bar Y=\frac{1}{n}\sum_{i=1}^n Y_i$
    1. First observation of sample: $Y_1$
    
## Coin toss example

- A _fair coin_ has equal probability of heads and tails
- How do we deterimine if a coin is fair?

> 1. Toss the coin many times 
> 1. Assign 1 if outcome is heads and -1 if tails 
> 1. Take the average
> 1. Check if the average is close to zero

> - But how can be sure it is fair if for example the average is 0.1 instead of zero?
>   - How sure are we?

# Tests

## Expected value and average

- What is the relationship between the expected value in the population and sample means taken from the population?
- Can we say how close the average height of 100 randomly sampled people is to the mean value in the population (the expected value)?
- Two approaches:
  - Calculate the distribution of the mean 
  - The Central Limit Theorem (CLT)

## Design of a test

- Select a _null hypothesis_ $H_0$
- Create a statistic with a known distribution under $H_0$
- Calculate the value of this statistic for the data
- Determine the probability of actual outcome (or worse) occuring assuming $H_0$
- Reject $H_0$ if the probability of the actual outcome occurring is below a chosen critical level

## Exact distribution

* If we know the distribution of the random variable $X$, we can calculate the distribution of $\bar X$
  * Binomial distribution of fair coin
  * Normal distribution
* Given this calculated distribution, we can see how likely the mean in our sample is
* Reject $H_0$ if it is _very_ unlikely (given some choice of threschhold)

## Coin toss average - distribution

```{r , echo=FALSE}
inputPanel(
  sliderInput("drawsb", label = "Draws:",
              min = 1, max = 50, value = 1, step = 1)
)
renderPlot({
    d = input$drawsb
x = -1 + 0:d * 2/ d
y = dbinom(0:d, d, 1/2)
ggplot(data.frame(x,y), aes(x, y)) +
    geom_bar(stat = "identity") +
    labs( x = bquote(bar(Y)), y = "probability")
})
```

## Almost normal distribution

```{r , echo=FALSE}
inputPanel(
  sliderInput("drawsa", label = "Draws:",
              min = 1, max = 50, value = 1, step = 1)
)
renderPlot({
    d = input$drawsa
x = -1 + 0:d * 2/ d
y = dbinom(0:d, d, 1/2)
ggplot(data.frame(x,y), aes(x, y)) +
    geom_bar(stat = "identity") +
    stat_function(fun=function(x){dnorm(x, sd=sqrt(1/d))*2/d}, color="red") +
    labs( x = bquote(bar(Y)), y = "probability")

})
```

## Central Limit Theorem (CLT)

- Properties of random samples

- Central Limit Theorem (CLT)
    - Take average of independent random draws
    - Draws can have (almost) any distribution
    - Large sample imply that estimates are almost normal

- Normal distributions are characterized by mean and variance
    - Thus focus is on the correctness of these statistics

- Bias for estimate
- Consistency (asymptotic bias)
- For CI and hypothesis tests variance estimate is important


## Tests

- Two types of tests
1. Hypothesis tests - test restrictions on the parameters
2. Specification tests - test that the model is correctly specified

## Errors

Two types of errors possible in hypothesis testing.

- _Type 1 error_: Reject $H_0$ but it is true
- _Type 2 error_: Do not reject $H_0$ but it is false

## Fair coin plot

 _p-value_ given by green areas
 
```{r , fig.height = 3}
prob = 0.05
avnorm = function(x){dnorm(x,0,1/sqrt(20))}
xval1 <- qnorm(prob/2,0,1/sqrt(20))
xval2 <- qnorm(1-prob/2,0,1/sqrt(20))
ggplot(data.frame(x=c(-1, 1)), aes(x)) +
    stat_function(fun= avnorm, geom="line") +
    geom_area(stat = "function", fun = avnorm, 
        fill = "green", xlim = c( -1, xval1), alpha=0.3) +
    geom_area(stat = "function", fun = avnorm, 
        fill = "green", xlim = c( xval2, 1), alpha=0.3) +
    labs(x="x", y="frequency", title="PDF") + 
    geom_vline(xintercept = 0, color="blue") + 
    geom_vline(xintercept = xval2, color="red") 

```

## Is a coin fair?

- Assign 1 to head and -1 to tail
- Under null hypothesis $H_0$ is fair $E(Y)=\mu_Y=0$ (blue line)
    - Variance is known!: $Var(Y)=1$
- With alternate hypothesis $H_1$ we have $E(Y)<0$ or $E(Y)>0$
- What is the probability that a sample average $\bar Y$ is as far away as $\hat Y$ or more?
    - Here estimate $\hat Y=0.4$ (red line)
    - If $\hat Y=0.4$ is possible, then also $\hat Y=-0.4$ should be
    - Called the _p-value_ (green areas)

## Coin toss average - find distribution

```{r}
shinyApp(
    ui = fluidPage(
        sliderInput("drawsd", label = "Draws:",
                    min = 10, max = 200, value = 10, step = 5),
        plotOutput("distPlot")
    ),
    server = function(input, output) {
        output$distPlot = renderPlot({
            d = input$drawsd
            x = -1 + 0:d * 2/ d
            y = dbinom(0:d, d, 1/2)
            ggplot(data.frame(x,y), aes(x, y)) +
                geom_bar(stat = "identity") +
                stat_function(fun=function(x){dnorm(x, sd=sqrt(1/input$drawsd))*2/input$drawsd}, color="red") +
                labs( x = bquote(bar(Y)), y = "frequency")
        })
    }
)
```

## Notation - population and sample

There is some more or less standard notation used in the book. If $Y$ 
and $Y$ are random variables

Concept | Population | Sample
------------------ | ---------------- | -----------
Expected value / Mean | $\mu_Y = E(Y)$ | $\bar Y$
Variance | $\sigma^2_Y = Var(Y)$ | $\hat\sigma^{2}_{Y} = s^{2}_{Y}$
Standard deviation / error | $\sigma_Y$ | $se(Y) = s_{Y} = \hat\sigma_{Y}$

- Called: 
    - Standard deviation in population
    - Standard error in sample

## Notation - population and sample

Population and sample definitions 

Concept | Population | Sample
------------------ | ---------------- | -----------
Expected value | $\mu_Y = E(Y)$ | $\bar Y = \frac{1}{n}\sum_i Y_i$
Variance | $\sigma^2_Y = E[(Y - \mu_Y)^2]$ | $s^{2}_{Y} = \frac{1}{n-1}\sum_i (Y_i - \bar Y)^2$
Standard error | $\sigma_Y$ | $s_{Y} = se(Y) = \sqrt{\frac{1}{n-1}\sum_i (Y_i - \bar Y)^2}$

# Asymptotic theory

## Central limit theorem

- Assume that $Y$ has mean $\mu_Y$ and variance $\sigma_Y$ 
- Then for large $n$ the sample average $\bar Y$ has almost a normal distribution 
$$N(\mu_Y, \sigma_Y^2/n)$$
- As the sample variance $s_Y^2$ converges to the population variance $\sigma_Y^2$, we know approximately the distribution of the sample mean for large $n$.

## Cumulative Density Function (CDF)

- From $X$ to cumulative probability

```{r , fig.height = 12, echo = FALSE}
shinyApp(
  ui = fluidPage(
      sidebarLayout(
          sidebarPanel(
              sliderInput("level", label = "x value:",
                  min = -4, max = 4, value = 0, step = .1)
          ),
          mainPanel(
              plotOutput("cdfPlot")
          )
      )     
  ),
  
  server = function(input, output) {
    output$cdfPlot <- renderPlot({
        xval <- input$level
        p1= ggplot(data.frame(x=c(-4, 4)), aes(x)) +
            stat_function(fun= pnorm, geom="line") +
            labs(x="", y="probability", title="CDF") + 
            geom_vline(xintercept = xval, color="red") + 
            geom_hline(yintercept = pnorm(xval), color="red")
        p2= ggplot(data.frame(x=c(-4, 4)), aes(x)) +
            stat_function(fun= dnorm, geom="line") +
            geom_area(stat = "function", fun = dnorm, 
                fill = "green", xlim = c( -4, xval), alpha=0.3) +
            labs(x="X", y="frequency", title="PDF") + 
            geom_vline(xintercept = xval, color="red")
        grid.arrange(p1, p2)
    })
  }
)
```

## One tail test - critical level 

- From precision goal to critical level

```{r , fig.height = 12}
shinyApp(
  ui = fluidPage(
      sidebarLayout(
          sidebarPanel(
              sliderInput("prob", label = "Probability:",
                  min = 0, max = .1, value = 0.05, step = .01)
          ),
          mainPanel(
              plotOutput("ciPlot")
          )
      )     
  ),
  
  server = function(input, output) {
    output$ciPlot <- renderPlot({
        xval <- qnorm(input$prob)
        p1= ggplot(data.frame(x=c(-4, 4)), aes(x)) +
            stat_function(fun= pnorm, geom="line") +
            labs(x="", y="probability", title="CDF") + 
            geom_vline(xintercept = xval, color="red") + 
            geom_hline(yintercept = pnorm(xval), color="red")
        p2= ggplot(data.frame(x=c(-4, 4)), aes(x)) +
            stat_function(fun= dnorm, geom="line") +
            geom_area(stat = "function", fun = dnorm, 
                fill = "green", xlim = c( -4, xval), alpha=0.3) +
            labs(x="x", y="frequency", title="PDF") + 
            geom_vline(xintercept = xval, color="red")
        grid.arrange(p1, p2)
    })
  }
)
```

## Two tail test - critical levels 

- Probabily divided equally to low and high outcomes

```{r , fig.height = 12}
shinyApp(
  ui = fluidPage(
      sidebarLayout(
          sidebarPanel(
              sliderInput("prob2", label = "Probability:",
                  min = 0, max = .1, value = 0.05, step = .01)
          ),
          mainPanel(
              plotOutput("ci2Plot")
          )
      )     
  ),
  
  server = function(input, output) {
    output$ci2Plot <- renderPlot({
        xval1 <- qnorm(input$prob2/2)
        xval2 <- qnorm(1-input$prob2/2)
        p1= ggplot(data.frame(x=c(-4, 4)), aes(x)) +
            stat_function(fun= pnorm, geom="line") +
            labs(x="", y="probability", title="CDF") + 
            geom_vline(xintercept = xval1, color="blue") + 
            geom_vline(xintercept = xval2, color="red") + 
            geom_hline(yintercept = input$prob2/2, color="blue") +
            geom_hline(yintercept = 1-input$prob2/2, color="red")
        p2= ggplot(data.frame(x=c(-4, 4)), aes(x)) +
            stat_function(fun= dnorm, geom="line") +
            geom_area(stat = "function", fun = dnorm, 
                fill = "green", xlim = c( -4, xval1), alpha=0.3) +
            geom_area(stat = "function", fun = dnorm, 
                fill = "green", xlim = c( xval2, 4), alpha=0.3) +
            labs(x="x", y="frequency", title="PDF") + 
            geom_vline(xintercept = xval1, color="blue") + 
            geom_vline(xintercept = xval2, color="red") 
        grid.arrange(p1, p2)
    })
  }
)
```

# Simulation / Monte Carlo

## Random draws

* The function `sample(x, size, replace = FALSE, prob = NULL)` generates random draws

```{r, echo=TRUE}
y = sample(c(-1,1), 50, replace = TRUE)
```

## Monte Carlo coin tosses

* Let $y$ be 50 tosses of a fair coin, with outcomes $y \in \{-1, 1\}$:

```{r, echo=TRUE}
y = sample(c(-1,1), 50, replace = TRUE)
```

* Use functions `mean` or `sum` to calculate average
```{r, echo=TRUE}
sum(y)/50
mean(y)
```


## Calculate mean, variance and standard errors

```{r, echo=TRUE}
ymean = sum(y)/50
ydiff = y - ymean
yvar = sum(ydiff^2)/49
ysd = sqrt(yvar)
ymean
yvar
ysd
```

## Mean and and variance 

If $Y$ has mean $\mu_Y$ and std. dev. $\sigma_Y$, then $t$ is approxaimately standard normal
$$t = \frac{Y - \mu_Y}{\sigma_Y}$$

```{r}
df <- tibble(y = 2 * rnorm(15)+5, dens=dnorm(y, mean=5, sd=2))
sidebarLayout(
mainPanel(
renderPlot({
    tibble(Y = -3:5) %>% 
        ggplot( aes(Y)) +
        stat_function(fun=dnorm, args=list(mean=input$muPop, sd=input$sigmaPop), color="red") +
        geom_vline(xintercept = 0) +
        ylab("")

})
),
    sidebarPanel(
  sliderInput("muPop", label = "Mean:",
              min = 0, max = 1.4, value = 1.4, step = .1),
  sliderInput("sigmaPop", label = "Standard deviation:",
              min = .2, max = 1, value = .7, step = .1)
)
)
```

## Normal CDF and inverse CDF in R 

One tail test at 5% level, lower bound

```{r, echo=TRUE}
qnorm(0.05)

# Going the other way
pnorm(-1.64) 
```

The probability of $-1.96 \le t \le 1.96$ is 0.95

```{r, echo=TRUE}
pnorm(-1.96) + 1 - pnorm(1.96)

```

## t-statistic with estimated variance

- Transform sample average to standard normal variable $t$: 
$$t=\frac{\bar Y - \mu_{\bar Y}}{\hat\sigma_{\bar Y}}$$
- See how far this variable is from 0. 
- This statistic is called the t-statistic or t-ratio

## t-distribution

- Distribution with uncertain variance
- If we know that Y has normal distribution
- With uncertain variance the confidence interval has different distribution

[Uncertain variance](https://xkcd.com/2110/)

## t-statistic with unknown variance

- True sample variance is usually not known. Estimate:
$$s_Y^2=\frac{1}{n-1} \sum_{i=1}^n (Y_i - \bar Y)^2$$
- Calculate the standard error of $\bar Y$:
$$SE(\bar Y) = \hat\sigma_{\bar Y} = s_Y / \sqrt n$$
- Calculate the t-ratio: 
$$t=\frac{\bar Y - \mu_Y}{SE(\bar Y)}$$
- Reject null hypothesis $H_0$ at 5 % level if $t>1.96$ 

## Calculating p-value

- Calculate t-ratio $\hat t$ for observed average $\hat Y$
- How probable is a larger t-ratio in absolute terms?
- The probability  $\hat p$ is given by $\hat p=2\Phi(\hat t)$, where $\Phi(t)$ is the cumulative distribution function of a standard normal distribution

## Confidence interval

- Consider all possible true values for the expected value $\mu_Y$ 
- _Confidence interval:_ all values that cannot be rejected at the chosen confidence level (usually 5 %)
- We want to find all $\mu$ with t-ratio $-1.96<t<1.96$:
$$-1.96<\frac{\bar Y - \mu}{se(\bar Y)}<1.96$$
- Solving for $\mu$ we get
$$\bar Y - 1.96*se(\bar Y)< \mu<\bar Y + 1.96*se(\bar Y)$$

## Important distributions

If $X$ and $Y$ are independent standard normal vars

- $e^X$ has *lognormal distribution*
- $X^2$ has $\chi^2$ *distribution*, 1 degree of freedom
    - $X^2+Y^2$ has $\chi^2$ *distribution*, 2 degrees of freedom
- $\frac{X}{Y^2}$ has *t distribution*, 1 degree of freedom
- $\frac{X^2}{Y^2}$ has *F distribution*, (1,1) degrees of freedom
- $\frac{X}{Y}$ has *Cauchy distribution*

## t distrubution

```{r , echo=FALSE , fig.width = 4}
inputPanel(
  selectInput("dft", label = "Degrees of freedom:",
              choices = c(5, 10, 20, 30, 40, 50), selected = 5)
)

renderPlot({
    ggplot(data.frame(x=c(0, 3)), aes(x)) +
        stat_function(fun = dnorm, geom="line") +
        stat_function(fun = function(x)(dt(x, df=as.numeric(input$dft))), geom="line") +
        labs(x="X", y="frequency") + theme(legend.title=element_blank()) +
        geom_vline(xintercept = 0) + 
        geom_hline(yintercept = 0)
})
```

## Chi

```{r , echo=FALSE , fig.width = 4}
inputPanel(
    sliderInput("level", label = "Critical level:",
        min = 0, max = 10, value = 0, step = .1),
  selectInput("df", label = "Degrees of freedom:",
              choices = 1:10, selected = 1)
)

renderPlot({
    ggplot(data.frame(x=c(0, 10)), aes(x)) +
        stat_function(fun= function(x)(dchisq(x, df=as.numeric(input$df))/as.numeric(input$df)), 
            geom="line") +
        geom_area(stat = "function", 
            fun = function(x)(dchisq(x, df=as.numeric(input$df))/as.numeric(input$df)), 
            fill = "green", xlim = c( input$level, 10), alpha=0.3) +
        labs(x="X", y="frequency") + theme(legend.title=element_blank()) +
        geom_vline(xintercept = 0) + 
        geom_hline(yintercept = 0)
})
```


